# Setup 
```{r}
library(tidyverse)
library(here)
set.seed(42) # ensuring that random sample results are reproducible
```

# Loading the data
```{r}
persuade <- read_csv(here("data", "raw", "persuade_2.0_human_scores_demo_id_github.csv"))

glimpse(persuade)
```

# Cleaning the data 
```{r}
# Checking how many of the rows have NAs for the columns important to this analysis 
persuade %>%
  summarise(
    n_total = n(),
    na_full_text = sum(is.na(full_text)),
    na_score = sum(is.na(holistic_essay_score)),
    na_prompt = sum(is.na(prompt_name)),
    na_econ_disadv = sum(is.na(economically_disadvantaged)),
    na_essay_id = sum(is.na(essay_id_comp))
  )
```

```{r}
# Filtering out NAs for economically_disadvantaged
persuade_clean <- persuade %>% filter(!is.na(economically_disadvantaged))

glimpse(persuade_clean)
```

# Filtering the data 
```{r}
# Seeing the distribution of econmically disdvantged status across different prompts 
persuade_clean %>% 
  count(prompt_name, economically_disadvantaged) %>%
  arrange(desc(n))
```

```{r}
# Checking the distribution of essay scores across the three most economically disadvantaged spread out prompts
score_check <- persuade_clean %>%
  filter(prompt_name %in% c("Driverless cars", "Exploring Venus", "Facial action coding system")) %>%
  filter(!is.na(holistic_essay_score)) %>%
  count(prompt_name, economically_disadvantaged, holistic_essay_score) %>%
  pivot_wider(names_from = holistic_essay_score, values_from = n, values_fill = 0)

print(score_check, width = Inf)
```

After looking at the economically_disadvantaged distribution across prompts and the essay score distributions across the top three well spread out prompts in terms of economically_disadvantage statuses, it seems that the facial action coding system prompt has the best spread of essay scores in both the economically disadvantaged and not economically disadvantaged groups. So, I chose to sample entries from the facial action coding system prompt only for this audit.  

```{r}
# Sampling entries from the facial action coding system prompt
chosen_prompt <- "Facial action coding system"

pool <- persuade_clean %>%
  filter(prompt_name == chosen_prompt) %>%
  filter(!is.na(holistic_essay_score))

pool %>%
  count(economically_disadvantaged, holistic_essay_score) %>%
  arrange(economically_disadvantaged, holistic_essay_score)
```

The essay scores within the facial action coding system prompt are not evenly distributed. Most of them cluster around the middle score range, leaving only a few at the extreme scores of 1 or 6. So, a plain random sample of 25 economically disadvantaged and 25 not economically disadvantaged essays would likely only contain medium-scoring essays. To capture the full range of essay quality in the sample, I group essays into three score bands: low = scores 1-2, mid = scores 3-4, and high = scores 5-6. I then sample proportionally from each band within each economic status group. 

```{r}
# Defining a function that samples essays proportionally across three score bands (Low, Mid, High) within a given economic status group
sample_group <- function(data, group_value, n_total = 25) {
  filtered <- data %>%
    filter(economically_disadvantaged == group_value) %>%
    mutate(score_band = case_when(
      holistic_essay_score <= 2 ~ "Low",
      holistic_essay_score <= 4 ~ "Mid",
      TRUE ~ "High"
    ))

  banded <- filtered %>%
    group_by(score_band) %>%
    group_modify(~ slice_sample(.x, n = min(nrow(.x), ceiling(n_total / 3)))) %>%
    ungroup()

  banded %>%
    slice_sample(n = min(n_total, nrow(banded)))
}

# Sampling 25 essays from each economic status group using the banded approach
disadv_sample     <- sample_group(pool, group_value = "Economically disadvantaged", n_total = 25)
non_disadv_sample <- sample_group(pool, group_value = "Not economically disadvantaged", n_total = 25)


# Checking the score distribution within each sampled group to confirm the bands actually produced a spread 
disadv_sample %>% count(holistic_essay_score)
non_disadv_sample %>% count(holistic_essay_score)
```

# Building and saving the final sample
```{r}
final_sample <- bind_rows(disadv_sample, non_disadv_sample) %>%
  select(essay_id_comp, full_text, source_text, assignment,
         holistic_essay_score, economically_disadvantaged,
         ell_status, race_ethnicity, student_disability_status, word_count) %>%
  mutate(llm_score = NA_real_,
         llm_rationale = NA_character_)

write_csv(final_sample, here("data", "processed", "persuade_sample_for_llm_scoring.csv"))
```

```{r}
final_sample %>%
  count(economically_disadvantaged, holistic_essay_score) %>%
  pivot_wider(names_from = holistic_essay_score, values_from = n, values_fill = 0)
```

The final sample for the audit contains 50 essays: 25 from economically disadvantaged students and 25 from economically non-disadvantaged students for the "Facial action coding system" prompt. Within each group, essays that span the holistic 1-6 score range are included with roughly equal representation acorss the low(score 1-2), mid (score 3-4), and high (score 5-6) bands used for sampling. This was done to ensure that the essays that the LLM is presented essays of varying quality across both economic status groups rather than average-scoring essays. 




