# Setup 
```{r}
library(tidyverse)
library(here)
set.seed(42) # ensuring that random sample results are reproducible
```

# Loading the data
```{r}
persuade <- read_csv(here("data", "raw", "persuade_2.0_human_scores_demo_id_github.csv"))

glimpse(persuade)
```

# Cleaning the data 
```{r}
# Checking how many of the rows have NAs for the columns important to this analysis 
persuade %>%
  summarise(
    n_total = n(),
    na_full_text = sum(is.na(full_text)),
    na_score = sum(is.na(holistic_essay_score)),
    na_prompt = sum(is.na(prompt_name)),
    na_econ_disadv = sum(is.na(economically_disadvantaged)),
    na_essay_id = sum(is.na(essay_id_comp))
  )
```

```{r}
# Filtering out NAs for economically_disadvantaged
persuade_clean <- persuade %>% filter(!is.na(economically_disadvantaged))

glimpse(persuade_clean)
```

# Filtering the data 
```{r}
# Seeing the distribution of econmically disdvantged status across different prompts 
persuade_clean %>% 
  count(prompt_name, economically_disadvantaged) %>%
  arrange(desc(n))
```

```{r}
# Checking the distribution of essay scores across the three most economically disadvantaged spread out prompts
score_check <- persuade_clean %>%
  filter(prompt_name %in% c("Driverless cars", "Exploring Venus", "Facial action coding system")) %>%
  filter(!is.na(holistic_essay_score)) %>%
  count(prompt_name, economically_disadvantaged, holistic_essay_score) %>%
  pivot_wider(names_from = holistic_essay_score, values_from = n, values_fill = 0)

print(score_check, width = Inf)
```

After looking at the economically_disadvantaged distribution across prompts and the essay score distributions across the top three well spread out prompts in terms of economically_disadvantage statuses, it seemed that the facial action coding system prompt has the best spread of essay scores in both the economically disadvantaged and not economically disadvantaged groups. However, this is a text-dependent prompt and the soruce_text field for thid prompt only contained the citation ('Making Mona Lisa Smile', by Nick D'Alto) rather than the original source article, which trained human raters would have had access to when scoring source-based evidence use. I then trie dto look for this full source text but was not able to find it. Thus, I instead checked the Independent-task prompts (which require no source material) for a similarly good score spread

```{r}
# Seeing which prompts are Independent (no source text needed) vs Source-dependent
persuade_clean %>%
  distinct(prompt_name, task) %>%
  arrange(task, prompt_name)

#Getting the list of Independent prompt names and  checking score spread across economic status for those specific prompts
independent_prompts <- persuade_clean %>%
  distinct(prompt_name, task) %>%
  filter(task == "Independent") %>%
  pull(prompt_name)

score_check_independent <- persuade_clean %>%
  filter(prompt_name %in% independent_prompts) %>%
  filter(!is.na(holistic_essay_score)) %>%
  count(prompt_name, economically_disadvantaged, holistic_essay_score) %>%
  pivot_wider(names_from = holistic_essay_score, values_from = n, values_fill = 0)

print(score_check_independent, width = Inf)
```

I chose "Distance learning" to be the independent-task prompt that the final sample focused on. Unlike the text-dependent prompt, independent-task prompts do not require an external source article. "Distance learning" also has the ost complete score spread of any independent-task prompt since both the economically disadvantaged and not economically disadvantaged groups contain a meaningful number of essays at every score level, 1 through 6. 

```{r}
# Sampling entries from the facial action coding system prompt
chosen_prompt <- "Distance learning"

pool <- persuade_clean %>%
  filter(prompt_name == chosen_prompt) %>%
  filter(!is.na(holistic_essay_score))

pool %>%
  count(economically_disadvantaged, holistic_essay_score) %>%
  arrange(economically_disadvantaged, holistic_essay_score)
```

The essay scores within the facial action coding system prompt are not evenly distributed. Most of them cluster around the middle score range, leaving only a few at the extreme scores of 1 or 6. So, a plain random sample of 25 economically disadvantaged and 25 not economically disadvantaged essays would likely only contain medium-scoring essays. To capture the full range of essay quality in the sample, I group essays into three score bands: low = scores 1-2, mid = scores 3-4, and high = scores 5-6. I then sample proportionally from each band within each economic status group. 

```{r}
# Defining a function that samples essays proportionally across three score bands (Low, Mid, High) within a given economic status group
sample_group <- function(data, group_value, n_total = 25) {
  filtered <- data %>%
    filter(economically_disadvantaged == group_value) %>%
    mutate(score_band = case_when(
      holistic_essay_score <= 2 ~ "Low",
      holistic_essay_score <= 4 ~ "Mid",
      TRUE ~ "High"
    ))

  banded <- filtered %>%
    group_by(score_band) %>%
    group_modify(~ slice_sample(.x, n = min(nrow(.x), ceiling(n_total / 3)))) %>%
    ungroup()

  banded %>%
    slice_sample(n = min(n_total, nrow(banded)))
}

# Sampling 25 essays from each economic status group using the banded approach
disadv_sample     <- sample_group(pool, group_value = "Economically disadvantaged", n_total = 25)
non_disadv_sample <- sample_group(pool, group_value = "Not economically disadvantaged", n_total = 25)


# Checking the score distribution within each sampled group to confirm the bands actually produced a spread 
disadv_sample %>% count(holistic_essay_score)
non_disadv_sample %>% count(holistic_essay_score)
```

The banded sample for the economically disadvantaged group includes essays scoring 2 through 6, but no essay scoring exactly 1. This is because only 4 of the 68 essays in the Low band (scores 1-2) for this group were scored a 1, so it is likely that none were randomly selected. The not economically disadvantaged group's sample includes essays across all scores, 1 through 6. Since the banded sampling approach samples proportionally from each band rather than guaranteeing every individual score is included, missing a single rare score value within a band is expected and does not affect the overall goal of capturing low, mid, and high scoring essays in both groups.

# Building and saving the final sample
```{r}
final_sample <- bind_rows(disadv_sample, non_disadv_sample) %>%
  select(essay_id_comp, full_text, holistic_essay_score,
         economically_disadvantaged, ell_status, race_ethnicity,
         student_disability_status, word_count) %>%
  mutate(llm_score = NA_real_,
         llm_rationale = NA_character_)

write_csv(final_sample, here("data", "processed", "persuade_sample_for_llm_scoring.csv"))
```

The final sample for the audit contains 50 essays: 25 from economically disadvantaged students and 25 from economically non-disadvantaged students for the "Distance learning" prompt. Within the not economically disadvantaged group, essays span the full 1-6 score range; within the economically disadvantaged group, essays span scores 2-6, with no essay scoring exactly 1 included (see note above). This ensures the LLM is presented with essays of varying quality in both economic status groups, rather than only average-scoring essays.





